Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-10-06 paper/review/response-to-review.Rmd

o The rule for capitalizing the starting letters of Figure, Section and Table is as follows: If you are referring to a particular figure/section/table then capitalize the first letter, otherwise use a lower-case first letter. For example, something shown in Section 4 vs. there are three sections in this paper.

> Appropriate instances of `section` and `figure` have been replaced with `Section` and `Figure`.

o All captions should appear below the corresponding figure/table. The captions should be in sentence style and end with a period.  No additional formatting (such as \emph, f or \it) should be used for the caption.

> Captions are below every figure, and for each table these have been moved manually in the LaTeX output.

@@

o All table row/column headers should also be in sentence style. There should not be further footnote-style annotations in tables; these should all be placed in the caption.

> Row/Column headers have been changed to sentence style, where appropriate. Some tables represent the direct output of a function, for example, in Table 2, 4, and 5. This is noted in the table caption that it is the output of a computation.

@@

o As a reminder, please make sure that:
	- \proglang, \pkg and \code have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.

> This has been done except for titles and references.

References:

o Journal of the Royal Statistical Society B (not: Journal of the Royal Statistical Society, Series B) the EM Algorithm.” 
o Springer-Verlag (not: Springer) and GGobi. 1st edition. Springer Publishing Company, Incorporated.

> These have been fixed.

o Please make sure that all software packages are \cite{}'d properly.

> These have been altered in the .bib file to include the proper \proglang{} and \pkg{} syntax.

o All references should be in title style.

> References are now in title style

o See FAQ for specific reference instructions.

@@

Code:

o Please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script.

> This has been provided in the "tidy-missing-data-paper.R" file

### 2021-05-19 paper/tidy-missing-data-paper.Rmd
Address reviewer comments

However, something must be known about the missing value structure to produce a complete dataset for analysis, whether by case- or variable-deletion, or with imputation. Decision tree models or latent group analysis can reveal structures or patterns of missingness [@Tierney2015; @Barnett2017], but are not definitive. While there are times where the missing data mechanism is obvious, determining this is typically not straightforward. To understand their structure and possible mechanisms, the analyst must explore the data with visualizations, summaries, and modelling, in an iterative fashion.

@@

The primary focus of most missing data packages is making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns. Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@vanBuuren2018]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

@@

@@

We propose a new form for tidy representation of missing data based on these ideas from past research. Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:nabularfig).

@@

3.  **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with "var_NA" being a random variable of the missingness of a variable, "var". This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.

4.  **Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, *nabular* data format, in sync with the data. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).

@@

*Nabular* data binds the shadow matrix column-wise to the original data. It is a portmanteau of "NA" and "tabular". *Nabular* data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)). *Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of "var", as "var_NA" during analysis. *Nabular* data is a snapshot of the missingness of the data. This means when *nabular* data are imputed, those imputed values can easily be identified in analysis ("var" vs "var_NA"). *Nabular* data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data. Firstly, missing value labels "NA" and "!NA" are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place. Imputing values on *nabular* data automatically tracks these imputations.

Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \@ref(verbs)).

@@

\pkg{naniar} does not reinvent the wheel for imputation, instead working with existing methods. However, \pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}. While useful to explore structure in missingness, they are not recommended for use in analysis. \code{impute_below()} imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.

@@

knitr::kable(list(t1, t2),
             caption = "Summary tables to help understand missingness patters. Output of \\code{miss\\_var\\_table(housing)} (left), tabulating missingness for variables, and output of \\code{miss\\_case\\_table(housing)}. There are 14 variables with 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings. About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
             digits = 1)
             # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

Other data structures such as spatial data, time series, networks, and longitudinal data would be supported by the inherently tabular, *nabular* data, if they are first structured as wide tidy format. Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation. Special missing value codes could be improved by creating special classes, or expanding low level representation of \code{NA} at the source code level.

>>>>

However, something must be known about the missing value structure to produce a complete dataset for analysis, whether by case- or variable-deletion, or with imputation. Decision tree models or latent group analysis can reveal structures or patterns of missingness [@Tierney2015; @Barnett2017], but are not definitive. While there are times where the missing data mechanism is obvious, determining this is typically not straightforward. A straightforward case is where missing values are encoded in the data in a non standard format that still indicates missingness, such as "N.A.", "N/A", or "missing". However, identifying these amongst all data can be problematic as it is hard to know what you are looking for, like finding a needle in a haystack. A more challenging case is where missing values arise due to the increasing or decreasing values of other variables (including themselves). For example, in weather data, temperature values might go missing due to their own high temperature. Or similarly, other measurements go missing due to high temperatures. Identifying this requires exploring the data, and the analysis path to take is obvious, as missing values are often omitted from plots and other statistical summaries. To understand their structure and possible mechanisms, the analyst must explore the data with visualizations, summaries, and modelling, in an iterative fashion.

@@

Missing data packages are focused primarily on making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns. Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@vanBuuren2018]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

@@

The \pkg{inspectdf} package provides tools for summarising, comparing, and visualising dataframes [@inspectdf]. In particular the `inspect_na()` function provides a summary of missing values for variables. `inspect_cat()` summarises the levels of categories in each column in a dataset. The `show_plot()` function shows a higher lever overview of the contents of a dataframe and its categories, as well as missing values. All of these functions allow for comparing missing values across two dataframes. For example, comparing the missing values in two dataframes with the same column names.

@@

We propose a new form for tidy representation of missing data based on these ideas from past research. Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:nabularfig). These are briefly described below, and discussed in further detail in section \@ref(nabular-data).

@@

3.  **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with "var_NA" being a random variable of the missingness of a variable, "var". This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state. Although this might break naming conventions for lower case "snake_case" variables (e.g., @Wickham2014), this distinguishes these columns from other columns.

4.  **Connectedness**: Binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).

@@

*Nabular* data binds the shadow matrix column-wise to the original data. It is a portmanteau of "NA" and "tabular". *Nabular* data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)). *Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of "var", as "var_NA" during analysis. *Nabular* data is a snapshot of the missingness of the data. This means when *nabular* data are imputed, those imputed values can easily be identified in analysis ("var" vs "var_NA"). Imputing values on *nabular* data automatically tracks these imputations. Together these features make it possible to do tasks that are not possible by simply adding a binary missing matrix, $B$ during data analysis.

@@

*Nabular* data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn]. Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \@ref(verbs)).

@@

\pkg{naniar} does not reinvent the wheel by creating the need for customised imputation procedures. Instead, the \code{nabular} format means existing methods using imputation replacement in place now automatically track missing values. \pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}. While useful to explore structure in missingness, they are not recommended for use in analysis. \code{impute_below()} imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.

@@

knitr::kable(list(t1, t2),
             caption = "Summary tables to help understand missingness patters. Output of \\code{miss\\_var\\_table(housing)} (left), tabulating missingness for variables, and output of \\code{miss\\_case\\_table(housing)}. There are 13 variables with 0-3 missings, 6 variables have 6000-10000 missings, 2 variables have 15000 – 17000 missings. About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
             digits = 1)
             # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

Other data structures such as spatial data, time series, networks, and longitudinal data would be supported by the inherently tabular, *nabular* data, if they are first structured as wide tidy format. Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation. Special missing value codes could be improved by creating special classes, or expanding low level representation of \code{NA} at the source code level. Similarly, list columns, present a challenge for analysis as their expansion into vectors in a dataframe is not always possible as they have different lengths, but should still be flagged for further analysis.
